From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-09-21 WOLFHPC15/article/article.tex
G5k ack
corrections article reviewer 3 + bib en plus Halide DSL -  Helene

%----------------------------------------
\begin{abstract}
  High performance architectures evolve continuously to be more powerful. Such architectures also usually become more difficult to use efficiently. As a scientist is not a low level and high performance programming expert, Domain Specific Languages (DSLs) are a promising solution to automatically and efficiently write high performance codes. However, if DSLs ease programming for scientists, maintainability and portability issues are deported from scientists to DSL designers. This paper deals with an approach to improve maintainability and programming productivity of DSLs through the generation of a component-based parallel runtime. To study it, the paper presents a  DSL for \emph{multi-stencil} programs, that is evaluated on a real-case of shallow water equations.
\end{abstract}
%----------------------------------------

@@

%----------------------------------------
\section{Introduction}
\label{sect:intro}
\input{introduction.tex}
%----------------------------------------
\section{Stencil and Multi-Stencil}
\label{sect:concept}
\input{concept.tex}
%----------------------------------------
\section{MSL and MSCAC Overview }
\label{sect:mscac}
\input{mscac.tex}
%----------------------------------------
\section{Language and Parallelization Description}
\label{sect:msmsc}
\input{ms_msc.tex}
%----------------------------------------
\section{A Component based Back-End}
\label{sect:component}
\input{component.tex}
%----------------------------------------
\section{Real-case Evaluation}
\label{sect:eval}
\input{evaluation.tex}
%----------------------------------------
\section{Related Work}
\label{sect:related}
\input{related.tex}
%----------------------------------------
\section{Conclusion}
\label{sect:conclusion}
\input{conclusion.tex}
%----------------------------------------

@@

We acknowledge Grid'5000 for evaluations.

>>>>

%----------------------------------------
\begin{abstract}
  High performance architectures evolve continuously to be more powerful. Such architectures also usually become more difficult to use efficiently. As a scientist is not a low level and high performance programming expert, Domain Specific Languages (DSLs) are a promising solution to automatically and efficiently write high performance codes. However, if DSLs ease programming for scientists, maintainability and portability issues are transferred from scientists to DSL designers. This paper deals with an approach to improve maintainability and programming productivity of DSLs through the generation of a component-based parallel runtime. To study it, the paper presents a  DSL for \emph{multi-stencil} programs, that is evaluated on a real-case of shallow water equations.
\end{abstract}
%----------------------------------------

@@

%----------------------------------------
\section{Introduction}
\label{sect:intro}
\input{introduction.tex}
%----------------------------------------
\section{Stencil and Multi-Stencil}
\label{sect:concept}
\input{concept.tex}
%----------------------------------------
\section{MSL and MSCAC Overview }
\label{sect:mscac}
\input{mscac.tex}
%----------------------------------------
\input{ms_msc.tex}
%----------------------------------------
\section{A Component based Back-End}
\label{sect:component}
\input{component.tex}
%----------------------------------------
\section{Real-case Evaluation}
\label{sect:eval}
\input{evaluation.tex}
%----------------------------------------
\section{Related Work}
\label{sect:related}
\input{related.tex}
%----------------------------------------
\section{Conclusion}
\label{sect:conclusion}
\input{conclusion.tex}
%----------------------------------------

@@

Experiments presented in this paper were carried out using the
Grid'5000 testbed, supported by a scientific interest group hosted by
Inria and including CNRS, RENATER and several Universities as well as
other organizations (see \url{https://www.grid5000.fr}).

### 2015-09-21 WOLFHPC15/article/concept.tex
corrections article reviewer 3 + bib en plus Halide DSL -  Helene

%To resume, concepts used in this paper to define a stencil kernel are \emph{mesh}, \emph{mesh entities}, \emph{data}, \emph{computation domain}, \emph{numerical expression} and \emph{stencil shape}. 
If stencil kernels have been studied a lot, the formalization of real overall mesh-based numerical simulations is a rather new contribution. Actually, paying attention to complex numerical simulations, it appears that most of them are composed of more than one stencil kernel, with one or more stencil shapes and of additional local computations. For example, how to formalize and parallelize a numerical simulation which chains the stencil kernels of Figures~\ref{fig:ex1} and~\ref{fig:ex2}. The next section formalizes concepts of \emph{stencil kernel}, \emph{local kernel} and \emph{multi-stencil program}.

%----------------------------------------
\subsection{A Multi-Stencil Formalization}
\label{sect:multistencil}
%----------------------------------------
This section introduces some formal definitions used in the rest of the paper.
Let $\Delta$ be the set of data of the simulation
A stencil kernel $s$ is defined as the quintuplet:
\begin{equation} 
s(R,w,exp,d),
\label{eq:st}
\end{equation}
where $R$ is a set of pairs $(r,n)$, with $r \in \Delta$ a data read by the computation, and $n$ the stencil shape (neighborhood) used to access $r$. The data written by the kernel is denoted $w \in \Delta$. $exp$ is the numerical expression of the stencil kernel. Finally the numerical expression is applied on the computation domain $d$.

>>>>

%To resume, concepts used in this paper to define a stencil kernel are \emph{mesh}, \emph{mesh entities}, \emph{data}, \emph{computation domain}, \emph{numerical expression} and \emph{stencil shape}. 
While stencil kernels have been studied a lot, the formalization of real overall mesh-based numerical simulations is a rather new contribution. Actually, paying attention to complex numerical simulations, it appears that most of them are composed of more than one stencil kernel, with one or more stencil shapes and of additional local computations. For example, how to formalize and parallelize a numerical simulation which chains the stencil kernels of Figures~\ref{fig:ex1} and~\ref{fig:ex2}. The next section formalizes concepts of \emph{stencil kernel}, \emph{local kernel} and \emph{multi-stencil program}.

%----------------------------------------
\subsection{A Multi-Stencil Formalization}
\label{sect:multistencil}
%----------------------------------------
This section introduces some formal definitions used in the rest of the paper.
Let $\Delta$ be the set of data of the simulation
A stencil kernel $s$ is defined as the quadruplet:
\begin{equation} 
s(R,w,exp,d),
\label{eq:st}
\end{equation}
where $R$ is a set of pairs $(r,n)$, with $r \in \Delta$ a data read by the computation, and $n$ the stencil shape (neighborhood) used to access $r$. The data written by the kernel is denoted $w \in \Delta$. $exp$ is the numerical expression of the stencil kernel. Finally the numerical expression is applied on the computation domain $d$.

### 2015-09-21 WOLFHPC15/article/evaluation.tex
corrections article reviewer 3 + bib en plus Halide DSL -  Helene

\begin{figure}[!h]
 \begin{center}
 \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
    \hline 
   Level & 1 & 2 & 3 & 4 & 6 & 10 & 12 & 16\\
   \hline
   Frequence & 2 & 1 & 3 & 5 & 3 & 1 & 1 & 2\\
   \hline
 \end{tabular}
\caption{Parallelism level (number of parallel tasks) and the number of times this level appears.}
\label{fig:freq}
 \end{center}
\end{figure}

@@

In addition to this, the scalability slope, shown by the number of iterations per second, is close to SkelGIS. This is an important result as the execution time of MSL is at the same time improved compared to SkelGIS. Actually, the less the execution time is, the less the scalability capacity is. Thus, these results confirm the result of~\cite{l2c} that using a component-based runtime does not damage performance of the back-end code. Finally, we notice that SkelGIS has proved its scalability on the Shallow water equations compared to an MPI parallelization~\cite{CPE:CPE3494}, that makes out performance results relevant.

>>>>

\begin{figure}[!h]
 \begin{center}
 \begin{tabular}{|c|c|c|c|c|c|c|c|c|}
    \hline 
   Level & 1 & 2 & 3 & 4 & 6 & 10 & 12 & 16\\
   \hline
   Frequency & 2 & 1 & 3 & 5 & 3 & 1 & 1 & 2\\
   \hline
 \end{tabular}
\caption{Parallelism level (number of parallel tasks) and the number of times this level appears.}
\label{fig:freq}
 \end{center}
\end{figure}

@@

In addition to this, the scalability slope, shown by the number of iterations per second, is close to SkelGIS. This is an important result as the execution time of MSL is at the same time improved compared to SkelGIS. Actually, the less the execution time is, the less the scalability capacity is. Thus, these results confirm the result of~\cite{l2c} that using a component-based runtime does not damage performance of the back-end code. Finally, we notice that SkelGIS has proved its scalability on the Shallow water equations compared to an MPI parallelization~\cite{CPE:CPE3494}, that makes our performance results relevant.

### 2015-09-21 WOLFHPC15/article/introduction.tex
corrections article reviewer 3 + bib en plus Halide DSL -  Helene

Many solutions to automatically optimize and parallelize numerical simulations have been proposed in the literature, either as specific frameworks~\cite{CPE:CPE3494,Trilinos-Overview,petsc-efficient} or as DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS}. Most of the time, DSLs focus on the parallelization and the optimization of a single numerical computation, also called a stencil kernel.
However, a real case numerical simulation (such as fluid dynamics, magneto-hydrodynamics, molecular dynamics etc.) is most of the time not composed of a single stencil kernel, but of a set of stencil kernels and a set of additional local auxiliary computations. The DSL proposed in this paper, called \emph{Multi-Stencil Language} or MSL, is a DSL to generate a parallel component-based structure of a numerical simulation, that we call a \emph{multi-stencil program}. As discussed in the related work, this work is complementary to the optimization and parallelization of a single stencil kernel.

>>>>

Many solutions to automatically optimize and parallelize numerical simulations have been proposed in the literature, either as specific frameworks~\cite{CPE:CPE3494,Trilinos-Overview,petsc-efficient} or as DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS, Ragan-Kelley:2013:HLC:2491956.2462176}. Most of the time, DSLs focus on the parallelization and the optimization of a single numerical computation, also called a stencil kernel.
However, a real case numerical simulation (such as fluid dynamics, magneto-hydrodynamics, molecular dynamics etc.) is most of the time not composed of a single stencil kernel, but of a set of stencil kernels and a set of additional local auxiliary computations. The DSL proposed in this paper, called \emph{Multi-Stencil Language} or MSL, is a DSL to generate a parallel component-based structure of a numerical simulation, that we call a \emph{multi-stencil program}. As discussed in the related work, this work is complementary to the optimization and parallelization of a single stencil kernel.
